Typed Functional Genetic Programming

### 2013-07-21 tex/thesis.tex
[tex]

\begin{todo}
& !!!!!!! zmínit $=_\beta$ a $\twoheadrightarrow_\beta$  a i pro ostatní redukce!!!!!
\end{todo}

@@

  \item[Logic point of view.] According to 
  the \textit{Curry–Howard correspondence}
  we can read statement $M : \sigma$ as 
  \textit{"$M$ is a proof of formula $\sigma$"}.
  We will not treat this interesting subject in much detail,
  \lets just say that $\ar$ corresponds to the
  implication $\then$ and \textit{atomic types} correspond to
  \textit{propositional variables}. The whole system of
  \textit{simply typed $\lambda$-calculus} then corresponds
  to the \textit{implicational fragment of propositional 
  intuitionistic logic}. We are mentioning this beautiful
  correspondence in hope that it will help clarify 
  use of \textit{inference rules} which are key concept
  in the process of deriving valid statements about
  terms and types.

@@

\newcommand{\bnf}{$\beta$-\textit{nf}\xspace}
\newcommand{\beenf}{$\beta\eta^{-1}$-\textit{nf}\xspace}

@@

The disadvantage of the \lnf, as the name suggests, is that it is long.
Terms in \lnf are said to be \textit{fully $\eta$-expanded}. 
Relevant property of $\eta$-reduction is that it always shortens the term
that is being reduced be it. And conversely, $\eta$-expansion prolongs.
$$\lamb{x}{(M~x)} \eArrow M \textbf{ ~~~~if } x \not\in FV(M) $$

Chceme ukazat že pomocí libovolné posloupnosti eta-redukcí se dostaneme do
unikátní eta-nf. To plyne z toho, že

 (a) každej lambda term má eta-NF (eta zmenšuje a nemuže se zmenšovat donekonečna)

 (b) eta-redukce je CR (z žlutýho barendrechta) a z toho plyne že má unikátní NF (lemátko)

pak chcem ukazat že eta redukcema nepřibude b-redex, když to dk
tak víme že sme se tema eta redukce dostali do betaeta-NF.

to se ukaže tak, že víme že eta redukce ničí @, tn ten novej redex by musel
mít svůj stare @. Zajímá nas levej "argument" @ , aby vznikl b-redex
musel by se zmenit něco co nebylo abstrakce na abstrakci ale to se nemohlo stát
protože se mohla tak akorat posunout blíž ta abstrakce M k tomu @
ale na zavinači určitě nevysel ten redukovanej podterm protože to by tam předtim měl ten redex.

@@

Now we will introduce \textit{Inhabitation tree}, structure slightly different from
\textit{Inhabitation machine}, which was introduced in \cite{barendregt10} by Barendregt.
We can think about Inhabitation tree as about unfolded Inhabitation machine.
The motivation for using Inhabitation trees is belief that it will help us 
reason about generation of \lterms 
of a given type $\sigma$ and with a given context $\Gamma$.

\subsection{Definition of Inhabitation tree}

\textit{Inhabitation tree} is a \textit{rooted tree}, possibly infinite. 
It has two types of nodes:

\begin{samepage}
\begin{itemize}
  \item Type nodes   
  			- containing type $\sigma \in \mathbb{T}$ 
  			- aka "OR-node" %, Nonterminal-node. 
  \item Symbol nodes 
  			%- containing $ x^{*} \in \mathfrak P \left(V\right) -blobost pač ordered $ or $c \in C$
  			- containing "$\lambda$-head" (nonempty finite sequence of variable names) or constant
  			  name. 
  			- aka "AND-node" %, Terminal-node. - ale nemusí, proto jsem to zakomentoval
\end{itemize}
\end{samepage}

We construct Inhabitation tree for given type $\sigma$ and context $\Gamma$.\\
We will define Inhabitation tree by describing its construction for a given $(\sigma,\Gamma)$.
Notice that it will closely follow the rules from \ref{barlike}:

\begin{itemize}
\item The root of Inhabitation tree for $(\sigma,\Gamma)$ is 
      \textit{type node} with $\sigma$ as type.
\item All \textit{type nodes} have as child nodes only \textit{symbol nodes}. 
\item And all \textit{symbol nodes} have as child nodes only \textit{type nodes}. 
\end{itemize}

\begin{samepage}
Now we will resolve the child nodes of the root node.
There are two cases of $\sigma$ (recall \ref{deftype}): 
\begin{description}
	\item[Atomic type] 
		$\sigma = \alpha $ where $\alpha \in A$. 
	\item[Function type] 
		$\sigma = \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$
		where $n \geq 1, \alpha \in A$.
\end{description}
\end{samepage}

First case \textbf{Atomic type} --- i.e., $\sigma = \alpha$ where $\alpha \in A$:\\
For every $(f,\rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha) \in \Gamma$
where $\alpha \in A$ there is a child \textit{symbol node} of the root containing constant name $f$.
This symbol node containing $f$ has $m$ child subtrees corresponding to Inhabitation trees for 
$(\rho_1,\Gamma),\dots,(\rho_m,\Gamma)$.

@@

Compare this case with corresponding grammar rule:
\[ 
	( \alpha , \Gamma )  \gar
	\bigg( \mbox{ {\Large f }}
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_1 , \Gamma )
	  \mbox{ \Vtextvisiblespace[1em] } 
	  \dots
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_m , \Gamma )
	  \quad \bigg)
\]

Second case \textbf{Function type} --- i.e., 
$\sigma = \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$
where $n~\geq~1, \alpha~\in~A$:\\
For every  $i \in \{1,\dots,n\}$ we create new \textit{variable name} $x_i$ which is not yet included in context $\Gamma$ as variable or constant name.

There is one and only one child \textit{symbol node} of the root containing "$\lambda$-head" 
$\lambda x_1 \dots x_n$ which stands for sequence of variable names $(x_1,\dots,x_n)$.
This symbol node containing $\lambda x_1 \dots x_n$
has one and only one child subtree corresponding to Inhabitation trees for 
$(\alpha,\Gamma \cup \{ (x_1,\tau_1) , \dots , (x_n,\tau_n) \})$.

@@

Compare this case with corresponding grammar rule:
\[ 
	( \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha , \Gamma )  \gar
	\bigg( \mbox{ {\Large 
	$\lambda x_1 \dots x_n .$ 
	}}( \alpha , \Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} ) \quad \bigg)
\]
~\\
\Lets consider following $(\sigma,\Gamma)$ as a simple example:

@@

\subsection{And-or tree and searching in Inhabitation tree}
\Lets consider following definition of 
\textit{And-or tree}\footnote{
\textbf{TODO}: Mention that on WIKI there is more general definition, but 
for our purposes is this one sufficient.}:

\textit{And-or tree} is a rooted tree where each node is labeled as either \textit{and-node} 
or\footnote{xor} \textit{or-node}.

By \textit{solving} And-or tree $T$ we mean finding finite tree $T'$
subgraph of $T$ such that it follows these conditions: 
\begin{itemize}
	\item The root of $T'$ is the root of $T$.
	\item Each \textit{and-node} in $T'$ has all the child nodes as in $T$.
	\item Each \textit{or-node}  in $T'$ has precisely one child 
	      node.\footnote{\textbf{TODO}: MENTION why precisely one 
	      and not at least one ..or CHANGE the def. }   
\end{itemize}
~
\Lets now consider following labeling of Inhabitation tree:

\begin{itemize}
  \item \textbf{Type nodes}   are labeled as \textbf{or-nodes}.   
  \item \textbf{Symbol nodes} are labeled as \textbf{and-nodes}.
\end{itemize}

This labeling has following justification:

Selection of exactly one
child node in \textit{type node} corresponds to selection of exactly one
grammar rule in order to rewrite nonterminal symbol.

Selection of all the child nodes in \textit{symbol node} corresponds to 
rewriting all the nonterminal symbols in string that is being generated.

@@

@@

\begin{todo}
  & DESCRIBE Inhabitation Machine...
   && něco jako na závěrještě ve stručnosti popišme 
      IM, strukturu ze který je odvozenej IT. 
  & říct, že je to hezkej konstrukt vzhledem k tomu že je konečnej
  & ale je popsanej velmi zrychla (v tý knize) a implementační
     detaily jsou celkem vynechaný
  & na stromový verzi jsou implementační detaily hezky vidět
  & a navíc pro polymorfní verzi jsou už i IM nekonečný takže
     se strácí i ta konečnostní výhoda
  & dát tam obrázek 
\end{todo}

>>>>

\begin{todo}
& !!!!!!! zmínit $=_\beta$ a $\twoheadrightarrow_\beta$  a i pro ostatní redukce!!!!!
& definovat \textbf{CR} a říct o tom lemátku, že má CR redukce má unikátní nf.
\end{todo}

@@

  \item[Logic point of view.] According to 
  the \textit{Curry–Howard correspondence}
  we can read statement $M : \sigma$ as 
  \textit{"$M$ is a proof of formula $\sigma$"}.
  We will not treat this interesting subject in much detail,
  \lets just say that $\ar$ corresponds to the
  implication $\then$ and \textit{atomic types} correspond to
  \textit{propositional variables}. The whole system of
  \textit{simply typed $\lambda$-calculus} then corresponds
  to the \textit{implicational fragment of propositional 
  intuitionistic logic}.

  Intuition is that the proof of $A \then B$ 
  can be imagined as function expecting proof of $A$
  and utilizing it in production of the proof of $B$
  that is returned as output of the function.

  We are mentioning this beautiful
  correspondence in hope that it will help clarify 
  use of \textit{inference rules}, which are key concept
  in the process of deriving valid statements about
  terms and types.

@@

\newcommand{\etar}{\twoheadrightarrow_\eta}
\newcommand{\ered}{$\eta$-reduction\xspace}
\newcommand{\bnf}{$\beta$-\textit{nf}\xspace}
\newcommand{\enf}{$\eta$-\textit{nf}\xspace}
\newcommand{\eenf}{$\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\beenf}{$\beta\eta^{-1}$-\textit{nf}\xspace}
\newcommand{\benf}{$\beta\eta$-\textit{nf}\xspace}
\newcommand{\bredex}{$\beta$-redex\xspace}

@@

The disadvantage of the \lnf, as the name suggests, is that it is long.
Terms in \lnf are said to be \textit{fully $\eta$-expanded} \cite{barendregt10}. 
Relevant property of $\eta$-reduction is that it always shortens the term
that is being reduced be it. And conversely, $\eta$-expansion prolongs.
$$\lamb{x}{(M~x)} \eArrow M \textbf{ ~~~~if } x \not\in FV(M) $$

Now we show that for every \lterm $M$ 
every sequence of $\eta$-reductions is finite and
leads to unique \enf $N$.

\begin{enumerate}
 \item Every application of \ered shortens the term.
       Since every term has finite size, this process must 
       end at some point. Thus every \lterm has \enf.
 \item Since \ered is \textbf{CR}, \enf is unique (see \ref{todo}). 
\end{enumerate}

So we can take every generated \lterm $M$ in 
\lnf and transform it to shorter term in \enf. 
The question is whether it remains in \bnf, thus being in \benf.
The answer is yes; it can be proven by showing that no 
new \bredex is created by \ered.

\begin{proposition}
Let $P$ be in \bnf and $P \eArrow Q$. Then $Q$ is in \bnf.    
\end{proposition}
\begin{proof}

For better clarity \lets show the \ered and \bredex using trees.\\
\ered: \\

\Tree [.$\lh{x}$ [.@ $M$ $x$ ] ] 
~~~~$\eArrow$~~~~
$M$

And \bredex:

\Tree [.@ [.$\lh{x}$ $M$ ] $N$ ] \\

\Lets assume that $P \eArrow Q$ creates a new \bredex $B$ in $Q$.

Since \ered only destroys and never creates \textit{function applications} (i.e. @),
the root @ of $B$ must be present in $P$.  
But since $P$ contains no \bredex, the left subterm $L$ of this root @
is not $\lambda$-abstraction.
Only possible way for $L$ to be changed by $\eArrow$ into 
a $\lambda$-abstraction is that $L$ is the reduced subterm (so that
$L$ is changed for its subterm).
But that is in contradiction with $P$ not containing any \bredex,
because it would cause $L$ be a $\lambda$-abstraction.

\end{proof}

Notable property of \lnf and \benf is that there is \textit{bijection} 
(i.e. one-to-one correspondence) of 
the set of simply typed \lterms in \lnf and 
the set of simply typed \lterms in \benf.

\begin{proposition}

Reduction to \enf is bijection between  
the set of simply typed \lterms in \lnf and 
the set of simply typed \lterms in \benf.
\end{proposition}
\begin{proof}

Since reduction to \enf always leads to an unique term, it is a function.
In previous proposition is shown that \ered of \lnf
leads to a term in \benf.\\

In order to show that a function is bijection it is sufficient to show that it is
both \textit{injection} and \textit{surjection}.\\

Suppose it is not injection.

So there must be $M_1,M_2$ in \lnf such that $M_1 \not= M_2$
and $N$ in \benf such that $M_1 \etar N$, $M_2 \etar N$.
Therefore $M_1 =_\eta M_2$, 
so $M_1 =_{\beta\eta^{-1}} M_2$.
This contradicts with $M_1,M_2$ being distinct \lnf{}s.\\

Every $M$ in \bnf has a \lnf $N$ such that 
$N \twoheadrightarrow_{\eta} M$ (proposition from \ref{lnf}).
Term $M$ in \benf is in \bnf, thus it has desired \lnf $N$
which reduces to it.

Therefore it is surjection. 
\end{proof}

Suppose we have systematic (i.e. gradually generating all terms, 
but no term twice) method for generating terms in \lnf,
we may transform it to systematic method for generating terms in \benf
by simply reducing each generated term to its \enf.

\begin{todo}
& možná eště zmínit něco jako : \\
Jedna drobná nevýhoda je, že pokud chceme generovat termy od nejmenšího do největšího
a pokud lnf generujeme od nejmenšího po největší tak pak generovaní \benf
timhle trikem nemusí bejt od nejmenšího po nějvětší. V naší aplikaci (tzn GP)
ale nijak nebazírtujeme na týhle vlastnosti a nakonec budeme spíš stejně používat
nějaký nedeterminizmy aby to fungovalo líp, takže to vubec nevadí, že budou trochu přeházený. 
\end{todo}

@@

Now we introduce \textit{inhabitation tree}; structure slightly different from
\textit{inhabitation machine}, which was introduced by Barendregt\cite{barendregt10}.
We can think about inhabitation tree as about unfolded inhabitation machine.
Inhabitation tree approach is further refinement 
of the term generating grammar approach, which is too much textual.
Whereas inhabitation tree approach aims at tree structure of \lterms.

~\\

\newcommand{\IT}{\mathop{\mathrm{IT}}}

Let context $\Gamma$ and type $\sigma$ be given.
\textit{Inhabitation tree} for $(\sigma,\Gamma)$, denoted as $\IT(\sigma,\Gamma)$,  
is a rooted possibly infinite tree that is recursively defined as follows.\\

If $\sigma = \alpha$ such that $\alpha \in A$, then

$\IT(\sigma,\Gamma) = $ \\

\Tree 
[.$\alpha$ 
   [.$f_1$ $\IT(\rho^1_1,\Gamma)$ $\dots$ $\IT(\rho^1_{m_1},\Gamma)$ ]
   [.$\dots$ ]
   [.$f_k$ $\IT(\rho^k_1,\Gamma)$ $\dots$ $\IT(\rho^k_{m_k},\Gamma)$ ]           
]\\[1em]

~~~~~~where

$~~~~~~~~~~\{ \,
f_1 : \rho^1_1 \ar \dots \ar \rho^1_{m_1} \ar \alpha ~,~\dots~,~
f_k : \rho^k_1 \ar \dots \ar \rho^k_{m_k} \ar \alpha \} \\ ~\,~~~~~~~~~~= 
\{\, f \: : \rho_1 \ar \dots \ar \rho_m \; \ar \alpha ~|~ 
  (f,\rho_1 \ar \dots \ar \rho_m \ar \alpha ) \in \Gamma \,\}$.\\

Otherwise, for $\sigma = \tau_1 \ar \dots \ar \tau_n \ar \alpha$ 
such that $n> 0, \alpha \in A$

$\IT(\sigma,\Gamma) =$\\

\Tree [.$\tau_1\ar\dots\ar\tau_n\ar\alpha$ 
 [.$\lh{x_1~\dots~x_n}$ [.$\IT(\alpha;\Gamma,x_1:\tau_1,\dots,x_n:\tau_n)$ ] ] ]

\end{definition}~\\[1em]

\Lets discuss this definition in grater detail. 
An inhabitation tree has two kinds of nodes.

\begin{samepage}
\begin{enumerate}
  \item \textit{Type node} (or \textit{OR-node}).

  	    This kind of node contains type. %$\sigma \in \mathbb{T}$  
  \item \textit{Symbol node} (or \textit{AND-node}).

  		This kind of node contains 
  		constant name, variable name 
  		\\or \textit{$\lambda$-head} 
  	    --- non-empty finite sequence of variable names. 
\end{enumerate}
\end{samepage}~

Following observation clarifies the alternation of node kinds.
\begin{enumerate}
\item The root of Inhabitation tree for $(\sigma,\Gamma)$ is 
      \textit{type node} with $\sigma$ as type.
\item All \textit{type nodes} have as child nodes only \textit{symbol nodes}. 
\item All \textit{symbol nodes} have as child nodes only \textit{type nodes}. 
\end{enumerate}~

%~\\~\\~\\

There are two cases of $\sigma$ --- 
\textit{atomic type} and \textit{function type}. \\

First case is \textbf{atomic type}, i.e., $\sigma = \alpha$ where $\alpha \in A$.

For every $(f,\rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha) \in \Gamma$
there is a child \textit{symbol node} containing constant or variable name $f$.
This symbol node containing $f$ has $m$ child subtrees corresponding to  
$\IT(\rho_1,\Gamma),\dots,\IT(\rho_m,\Gamma)$.

Therefore specially for $m = 0$ node containing $f$ is 
a leaf of the inhabitation tree.

@@

Compare this case with corresponding grammar rule.
\[ 
	( \alpha , \Gamma )  \gar
	\bigg( \mbox{ {\Large f }}
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_1 , \Gamma )
	  \mbox{ \Vtextvisiblespace[1em] } 
	  \dots
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_m , \Gamma )
	  \quad \bigg)
\]

Second case is \textbf{function type}, i.e., 
$\sigma = \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$
where $n~\geq~1, \alpha~\in~A$.\\
For every  $i \in \{1,\dots,n\}$ we create new \textit{variable name} $x_i$ 
that is not yet included in context $\Gamma$ as variable or constant name.

There is one and only one child \textit{symbol node} 
of function type node containing $\lambda$-head
$\lambda x_1 \dots x_n$, which stands for 
sequence of variable names $(x_1,\dots,x_n)$.
This symbol node containing $\lambda x_1 \dots x_n$
has one and only one child subtree corresponding to  
$\IT(\alpha;\Gamma,x_1:\tau_1 , \dots , x_n:\tau_n )$.

@@

Compare this case with corresponding grammar rule.
\[ 
	( \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha , \Gamma )  \gar
	\bigg( \mbox{ {\Large 
	$\lambda x_1 \dots x_n .$ 
	}}( \alpha ; \Gamma , x_1:\tau_1,\dots,x_n:\tau_n ) \quad \bigg)
\]
~\\
\Lets consider following $(\sigma,\Gamma)$ as a simple example:

@@

\newpage
\subsection{And-or tree and searching in Inhabitation tree}

\Lets consider following definition of 
\textit{and-or tree}\footnote{
\textbf{TODO}: Mention that on [WIKI] there is more general definition, but 
for our purposes is this one sufficient.}.

\begin{definition}
~
	\begin{enumerate}
	\item \textit{And-or tree} is a rooted possibly infinite 
	      tree with every node labeled as either \textit{and-node} 
	      or
	      %\footnote{xor} 
	      \textit{or-node}.

	\item By \textit{solving and-or tree} $T$ is meant finding finite tree $T'$
	      subgraph of $T$ such that following conditions are met. 
	    \begin{enumerate}
		  \item The root of $T'$ is the root of $T$.
		  \item Each \textit{and-node} in $T'$ has all the child nodes as in $T$.
		  \item Each \textit{or-node}  in $T'$ has precisely one child 
		      node.\footnote{\textbf{TODO}: MENTION why precisely one 
		      and not at least one ..or CHANGE the def. }   
	    \end{enumerate}

	\end{enumerate}

~\\[1em]
\Lets now consider following labeling of Inhabitation tree.

\begin{itemize}
  \item \textbf{Type nodes}   are labeled as \textbf{or-nodes}.   
  \item \textbf{Symbol nodes} are labeled as \textbf{and-nodes}.
\end{itemize}~

This labeling has following justification.\\

Selection of exactly one
child node in \textit{type node} corresponds to selection of exactly one
grammar rule in order to rewrite nonterminal symbol.  \\

Selection of all the child nodes in \textit{symbol node} corresponds to 
rewriting all the nonterminal symbols in string that is being generated.
Or put in different words, it corresponds to filling all the slots
for arguments with values.\\

@@

\begin{todo}
 & Přepsat poslední odstavec --- počítal ještě s tím že 
   pasaže o stromové reprezentaci jsou až po této kapitole, což bylo změněno.
\end{todo}

@@

\begin{todo}
  & Když na to zbude čas, tak popsat Inhabitation Machine...
   && něco jako na závěr ještě ve stručnosti popišme 
      IM, strukturu ze který je odvozenej IT. 
  & říct, že je to hezkej konstrukt vzhledem k tomu že je konečnej
  & ale je popsanej velmi zrychla (v tý knize) a implementační
     detaily jsou celkem vynechaný
  & na stromový verzi jsou implementační detaily hezky vidět
  & a navíc pro polymorfní verzi jsou už i IM nekonečný takže
     se strácí i ta konečnostní výhoda
  & dát tam obrázek 
\end{todo}
